---
title: "DS250 - Introduction to Data Science"
author: "Student Name"
date: "October 20, 2015"
output:  html_document
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```
```{r}
rm( list = ls())  # Clear environment
```

# Assignment 3 - Getting Started with Statistics

This assignment covers core concepts in data and descriptive and inferential statistics. Statistical analysis is one of the primary tools of a Data Scientist used in understanding data, testing validity, and making predictions.

### The objectives of this assignment:
* Understand the types of data
* Review core concepts in descriptive statistics
* High level review of probability distributions
* Practice making predictions using regression and probability

***
#### Question 1 - Types of Data
**a**\. Describe some of the differences between qualitative and quantitative data:

````

````

**b**\. Describe some of the differences between an interval and a ratio:

````

````

**c**\. From the image below, identify the statistical data types in the departure table:
(https://canvas.uw.edu/users/3264551/files/38442830/download?download_frd=1)


***
#### Question 2 - Distributions
**a**\. Central Tendency: Briefly describe the following measures:

* Mode
* Median
* Mean

````

````

**b**\. Dispersion: Briefly describe the following measures:

* Range
* Variance
* Standard deviation

````

````

**c**\. Using the following sequence, create a plot of the normal distribution with a mean = 0 and standard deviation = 1.

````{r echo=TRUE}
x <- seq(-5,3, 0.01)
````

***
#### Question 3 - Probabilities
**a**\. Using the heart patient data in the 'processed.cleveland.data.txt' file: (link below)   
(https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/heart-disease.names)  
(https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data)  

- separate the population into male and female groups  
- calculate the mean and standard deviation for male cholesterol values  
- create a histogram showing the distribution of male cholesterol values 

````

````

**b**\. Using **rnorm()**, create a normal distribution and compare the histogram of the male cholesterol against the simulated normal distribution. Are they the same? *(can also use dnorm())*

````

````

**c**\. Using **pnorm()**, determine the probability that a randomly chosen adult male has a cholesterol level of over 260.

````

````

***
#### Question 4 - Comparison
**a**\. Load sample data file from: (http://www.openintro.org/stat/data/cdc.R)  
*(use "source(file)" function to load external R data.)*

````

````

**b**\. Create a histogram plot of the weight values:

````

````

**c**\. Create a new data frame (*young*) from the cdc data file containing individuals between the ages of (18 to 30).

````

````

**d**\. Use **sample()**, to collect 50 random sample values from the (young) data frame.

````

````

**e**\. Describe the distribution of this sample?  
How does it compare to the distribution of the population from which it was sampled from?

````

````

**f**\. Using **sample()**, create three separate sample collections for (10, 50, 100) samples and plot the histogram of these three data sets. How are they different?  

*using the original cdc data, not the filtered 'young' data*

````

````

***
#### Question 5 - Confidence Intervals  
Using the sample data from: (http://www.openintro.org/stat/data/cdc.R)  
*(use "source()" to load an external file)*


**a**\. Create a data frame with the height values from the cdc sample data.

````

````

**b**\. Create a sample data frame with 60 random samples of individual height values. Based on this sample data, what it the average height for these 60 individuals?

````

````

**c**\. Calculate a confidence interval (upper, lower bounds) for the sample data.

````

````

**d**\. Write a short loop to create (50) separate sample sets and calculate the confidence intervals for each set. How do the confidence intervals change in the 50 samples?


````

````

***
#### Question 6 - Inference  
Use **"load()"** to import sample data file from: (http://www.openintro.org/stat/data/nc.RData) and complete the following:
    
**a**\. Review the attributes in this data set and determine which ones are categorical or numeric.

````

````

**b**\. Create a side-by-side boxplot of habit and weight. What does the plot highlight about the relationship between these two variables?

````

````

**c**\. Using the following statement to compare weight and habit. Is the difference between the mean of nonsmoker and that of smoker statistically significant?

````
by(nc$weight, nc$habit, mean)

````

**d**\. Use the following statement to calculate the inference measure between weight and habit.

````
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ht", null = 0, alternative = "twosided", method = "theoretical")

````

**e**\. In the statement above, change the type argument to 'ci' to construct a confidence interval for the difference between the weights of babies born to smoking and non-smoking mothers.

````

````

***
#### Question 7 - Linear Regression  
Load the sample data file from: (http://www.openintro.org/stat/data/mlb11.RData) and complete the following:

**a**\. Use the **cor()** function to determine the numeric relationship between 'runs' and 'at_bats'.

````

````

**b**\. Use the custom function **plot_ss()** *(included in RData file)* to create a plot for:

````
plot_ss(x = mlb11$at_bats, y = mlb11$runs, showSquares = TRUE)

````

**c**\. Review the mlb11 data set and select another variable that might be a good predictor of runs.

````{r echo=TRUE}

````

**d**\. Create a scatterplot of the two variables and fit a linear model. Determine if there is a linear relationship?

````{r echo=TRUE}

````

***
#### Question 8 - Linear Regression  
Use the built-in dataset (*faithful*) and complete the following:

a) Create a linear model *lm()* and capture the standard residuals :

````{r echo=TRUE}

````

b) Using *qqnorm()* create a normal probability plot with *qqline()* showing the comparison:

````{r echo=TRUE}

````

***
#### Question 9 - Estimates with Monte Carlo 

Review and correct the following code to obtain an approximate value of Pi

````{r echo=TRUE}
runs <- 10
#runif samples from a uniform distribution
xs <- runif(runs,min=-0.5,max=0.5)
ys <- runif(runs,min=-0.5,max=0.5)

in.circle <- xs^2 + ys^2 <= 0.5^2
mc.pi <- (sum(in.circle)/runs)*4

plot(xs,ys,pch='.',col=ifelse(in.circle,"blue","grey")
     ,xlab='',ylab='',asp=1,
     main=paste("MC Approximation of Pi =",mc.pi))
````

***
#### Question 10 (Optional) - Probability Puzzle 

In a country in which people only want boys, every family continues to have children until they have a boy. If they have a girl, they have another child. If they have a boy, they have no more children. What is the proportion of boys to girls in the country? (Explain your answer)

````


````

#### **End**
