---
  title: 'Titantic Survival Exploration'
  author: "Amy Werner-Allen"
  output: html_document

---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

```{r echo=FALSE}
library(reshape)
library(stringr)
library(ggplot2)
library(corrplot)
```
  
### Data Exploration

First and foremost, we need to take a peek at the data. There are two datasets on the [Kaggle](https://www.kaggle.com/c/titanic) competition site: a training data set and a test data set. I chose to combine both to get a sense of what the data looks like. 

```{r}
setwd("~/Documents/Repositories/datascience/Kaggle/Titantic")
titanic.train = read.csv("train.csv", header = TRUE, stringsAsFactors = FALSE)
titanic.train$IsTrain = TRUE
titanic.test = read.csv("test.csv", header = TRUE, stringsAsFactors = FALSE)
titanic.test$IsTrain = FALSE

titanic.test$Survived = NA
titanic = rbind(titanic.train, titanic.test)
```

So what do we have?

```{r}
str(titanic)
```

We have *1309* data points: 891 training data points with survived in (0, 1) and 418 test data points with unknown survival outcomes (NA).

We have 13 variables: <br>

* **Passenger ID**: a primary key to identify a passenger
* **Survived**: indicator of surivival (will be NA when IsTrain is FALSE)
* **Class**: the class designation (1st, 2nd, or 3rd)
* **Name**: name of passenger
* **Sex**: sex of passenger
* **Age**: numeric age of passenger (decimal if less than 1 years old)
* **Sibling & Spouse**: number of siblings or spouses also aboard
* **Parent & Child**: number of parents or children also aboard
* **Ticket**: the ticket number
* **Fare**: how much the ticket cost
* **Cabin**: the cabin number of the passenger
* **Embarked**: the port of embarkment 
* **IsTrain**: indicates whether data is from training or test dataset

Now let's take a look at some key features to see exactly what we have.

Starting with **age**, here is a quick distribution of the ages that we have for passengers:
```{r}
ggplot(titanic, aes(x=Age)) +
  geom_histogram(colour="darkblue", fill="#6699FF") +
  ggtitle("Age Distribution on Titanic") + 
  scale_fill_gradient(low = "light blue", high = "orchid4", guide=FALSE) +
  theme(panel.grid.minor = element_line(colour = "red", linetype = "dotted"), panel.background = element_rect(colour = "darkblue"))
```

About 20% of rows have no age information:
```{r}
sum(is.na(titanic$Age))
```
The same is not true for sex: all passengers have gender information filled out.

What about **survival rates**? Looking globally, the chance of survival is around 40%:
```{r}
table(titanic$Survived)
```

We can also look at survival rates by different factors. Here we have survival rate by sex:
```{r}
ggplot(titanic, aes(Survived, fill=Sex)) + 
  geom_bar(width = 0.8) +
  scale_fill_manual(values=c("#CC3399", "#6699FF")) +
  theme(panel.grid.minor=element_blank(),panel.grid.major=element_blank()) +
  scale_x_continuous(breaks=c(0,1), labels=c("No", "Yes"))
```

We can see that women had a higher chance of surviving than men.

Here is survival rate by class:
```{r}
titanic$Pclass = as.factor(titanic$Pclass)

ggplot(titanic, aes(Survived, fill=Pclass)) + 
  geom_bar(width = 0.8) +
  scale_fill_manual(values=c("#CC3399", "#6699FF", "#66CC33")) +
  theme(panel.grid.minor=element_blank(),panel.grid.major=element_blank()) +
  scale_x_continuous(breaks=c(0,1), labels=c("No", "Yes"))
```

Being in first class gave you a better chance of survival, while being in third class meant you were much more likely *not* to survive. 

How about age? Lets create a factor for age based on numeric years.

```{r}
titanic$AgeF = "Infant"
titanic$AgeF[titanic$Age >=2] <- "Child"
titanic$AgeF[titanic$Age >=18] <- "Adult"
titanic$AgeF[titanic$Age >=50] <- "Elderly"

ggplot(titanic[!is.na(titanic$Age), ], aes(Survived, fill=AgeF)) + 
  geom_bar(width = 0.8) +
  scale_fill_manual(values=c("#CC3399", "#66CCCC", "#CC99FF", "#66CC33")) +
  theme(panel.grid.minor=element_blank(),panel.grid.major=element_blank()) +
  scale_x_continuous(breaks=c(0,1), labels=c("No", "Yes"))
```

We can see that infants had a higher chance of surviving, whereas the elderly had a lower chance. 

What about the **fare**, how much passengers spent on their tickets?

```{r}
ggplot(titanic, aes(x=Fare)) +
  geom_histogram(colour="darkblue", fill="#6699FF", binwidth=10) +
  ggtitle("Fare Distribution on Titanic") + 
  scale_fill_gradient(low = "light blue", high = "orchid4", guide=FALSE) +
  theme(panel.grid.minor = element_line(colour = "red", linetype = "dotted"), panel.background = element_rect(colour = "darkblue"))
```

We can see that there is a long tail on this distribution, with some passengers spending more than $500 on their ticket. The average price of a ticket was $33.

### Feature Selection

Now that we have an idea about the data completeness and distributions, let's start looking at correlations between different data elements.

```{r}
titanic$Pclass = as.numeric(titanic$Pclass)
titanic$SexN = 1
titanic$SexN[titanic$Sex=="male"] = 0
# 0 is male, 1 is female

titanic_numeric = titanic[, sapply(titanic, is.numeric)]
titanic_numeric = titanic_numeric[,-1]
M = cor(titanic_numeric, use="pairwise.complete.obs")
corrplot(M, method="shade")
```
